---
title: "Simulating spatial-temporal data"
author: "`r Sys.Date()`"
output: 
  html_document:
    self_contained: yes
    number_sections: true
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    font: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(1019)

library(mgcv)
library(nlme)
library(tidyverse)
library(plotly)
library(mvtnorm)
library(ggpubr)
```

# Potential simulation scheme

The simulation, technically, would be established on a continuous space-time domain, even though the final "observations" would be a discrete realization.

Take a Gaussian process for example:

1. The observation is composed of a true latent process and an error process.

\[\begin{aligned}Y_i(\mathbf{s}, t) &= \eta_i(\mathbf{s}, t) +\epsilon_i(\mathbf{s}, t) \\
\epsilon_i(\mathbf{s}, t) & \sim GP(0, \Gamma_{\epsilon})\end{aligned}\]

- $\Gamma_{\epsilon}$ constant across all observations 
- Usually, realization of $\Gamma_{\epsilon}$ on any discrete grid would be $\sigma_{\epsilon}^2\mathbf{I}$. But I guess correlation of the error can be introduced by altering its structure? This is probably not what we are considering so far. 

2. The true latent process is composed of a fixed process and a random (subject-specific) process. 

\[\begin{aligned}
\eta_i(\mathbf{s}, t) &= \mu(\mathbf{s}, t)+b_i(\mathbf{s}, t) \\
b_i(\mathbf{s}, t) & \sim GP(0, \Gamma_{b})
\end{aligned}\]

- $\mu(\mathbf{s}, t)$ is the population mean, constant across subjects
- $b_i(\mathbf{s}, t)$ is the individual-level random effect
- We can introduce correlation between space/time by altering the structure of $\Gamma_{b}$

3. The fixed process can be a linear combination of covariates/basis functions

$$\mu(\mathbf{s}, t) = \sum_{p=1}^PX_p(\mathbf{s}, t)\beta_p$$

- $X(\mathbf{s}, t)$ can be either covariates or basis functions

4. The random process, $b_i(\mathbf{s}, t)$, can also be decomposed into the linear combination of basis functions, just like what we did in the fGFPCA project

\[\begin{aligned}b_i(\mathbf{s}, t) &= \sum_{k=1}^K\phi_k(\mathbf{s}, t)\xi_{ik} \\
\mathbf{\xi}_i& \sim N(0, \Gamma_{\xi})
\end{aligned}\]

This is a good way to incorporate complicated spatial-temporal correlation. We may be able to incorporate spatial-specific or temporal specific correlation by factorizing $\phi(s, t)$ in to $\phi(s)$ and $\phi(t)$ and manipulating the structure of $\Gamma_{\xi}$. 

Another way I can think of it is directly generating $b_i$ directly from a correlation matrix (e.g. AR1). But it seems to be it would be difficult to control on each domain. Also, the space-time domain is probably greater than 2D, on which a correlation matrix would be hard to envision. 


# A toy exmaple

Built up from last time, I would like to generate a dataset as follows:

- Sample size N = 100
- Temporal grid T = 10
- Spatial grid: two dimensional $\mathbf{s} = (s_1, s_2)$, where $s1, s2 \in [1, 2, ..., 10]$. I will hereafter denote the collection of all $\mathbf{s}$ as $S$, and the number of spatial grid points as $S_n$. 
- Assume for now data is collected on a regular grid


```{r}
# Sample size
N <- 100

# generate space grid
nS <- 100
s1 <- 1:10
s2 <- 1:10

# generate time grid
nT <- 10
t <- seq(0, 1, length.out = nT) 

# put in a long-format dataframe
df <- expand.grid(id = 1:N,
                  s1 = s1, s2= s2, 
                  t=t)
df <- df %>% arrange(id)
```

This means for each subject, there are 1000 measurements. 


- Assume the fixed process is constant zero

```{r}
# Fix effect
mu_func <- function(s1, s2, t){0}
df <- df %>% mutate(mu = mu_func(s1, s2, t))
```


## Generate random process

I am starting with a very simple, factorized random effect: 

\[\begin{aligned}
b_i(\mathbf{s}, t) &= \xi_{i1}\phi_1(\mathbf{s})+\xi_{i2}\phi_2(t)\\
\mathbf{\xi}_i & \sim N(0, \Gamma_{\xi}),\ 
\Gamma_{xi}  = \begin{bmatrix} 5 & 0 \\ 0 & 5 \end{bmatrix}
\end{aligned}\]

I would like to set up the spatial distribution such that it has a peak with the highest value, and the value decrease as it gets further from the center. In other words, the value of $\phi_1(\mathbf{s})$ decrease as the distance to $(5, 5)$ increases. 

On top of that, we also wish $\phi_1(\mathbf{s})$ to be bounded, and relatively smooth. 

The following function satisfies the conditions: 

$$\phi_1(\mathbf{s}) = \frac{1}{exp[\frac{(s1-5)^2+(s2-5)^2}{10}]}$$

```{r}
df <- df %>%
  mutate(phi1 = 1/exp(((s1-5)^2+(s2-5)^2)/10))
# range(df$phi1)
plot_ly(data.frame(df, s1=s1, s2=s2), x=~s1, y=~s2, z=~phi1) %>% add_markers(size = 5)
```

For temporal correlation, I will use a square root function:

$$\phi_2(t) = \sqrt{t}$$

```{r}
df <- df %>% 
  mutate(phi2 = sqrt(t))
# range(df$phi2)
```

Now we can generate scores and calculate the random effect. 

```{r}
xi <- matrix(rnorm(N*2, mean = 0, sd=5),
             nrow = N, ncol = 2)

# random process
df$bi <- NA
for(i in 1:N){
  df$bi[df$id==i] <- as.matrix(df[df$id==1, c("phi1", "phi2")]) %*% xi[i, ]
}
```

```{r}
df <- df %>%
  mutate(Y = mu+bi)

df$tid <- factor(df$t, levels = t, labels = 1:nT)
```

For simplicity, we for now do not consider random error $\epsilon$. 

```{r, eval=FALSE}
# Visualize one subject
df %>% 
  filter(id == 50) %>%
  ggplot()+
  geom_tile(aes(x=s1, y=s2, fill=Y))+
  facet_wrap(~tid)+
  labs(title="Time-startified plot for one subject (id=50)")

plot_ly(df %>% filter(id == 50), 
        x = ~s1, y = ~s2, z = ~ Y, color = ~tid) %>% add_markers(size = 5)

# xi[50, ]
```

# Empirial estimators

Here the data is three dimensional ($s_1, s_2, t$). It would be difficult to visualize, especially covariance matrix, since it is pariwise between every pairs of combination. In the textbook I was reading, mean and covariance at one dimention are calculated averaging over other dimensions. Details as below

## Empirical mean

Let's start with marginal mean. For example, empitical spatial mean is averaged over time

$$\bar{Y}(\mathbf{s}) = \frac{1}{NT}\sum_{i=1}^N \sum_{t=1}^T Y_i(\mathbf{s}, t)$$

```{r, warning=FALSE, message=FALSE}
df %>% 
  group_by(s1, s2) %>%
  summarize(emp_sp_mean = mean(Y)) %>%
  ggplot()+
  geom_tile(aes(x=s1, y=s2, fill=emp_sp_mean))+
  labs(title = "Empirical spatial mean")
```
Likewise, empirical temporal mean would marginalize over space:

$$\bar{Y}(t) = \frac{1}{NS_n}\sum_{i=1}^N \sum_{\mathbf{s} \in S} Y_i(\mathbf{s}, t)$$


```{r, warning=FALSE, message=FALSE}
df %>% 
  group_by(t) %>%
  summarize(emp_t_mean = mean(Y)) %>%
  ggplot()+
  geom_line(aes(x=t, y=emp_t_mean))+
  labs(title = "Empirical temporal mean")
```

I guess in the mean case, I can barely visualize the 3D funciton: 


$$\bar{Y}(s_1, s_2, t) = \frac{1}{N} \sum_{i=1}^N Y_i(s_1, s_2, t)$$

```{r, warning=FALSE, message=FALSE}
df %>% group_by(s1, s2, t) %>%
  summarise(emp_mean = mean(Y)) %>% 
  mutate(t=as.factor(round(t, 2))) %>%
  plot_ly(., x=~s1, y=~s2, z=~emp_mean, color=~t) %>%
  add_markers(size = 0.2)
  
```

## Empirical covariance

For this kind of spatial-temporal data, covariance/correlation matrix is a more complicated issue. Technically, it can be calculated between any pair of combination of space and time: 

$$Cov(Y(s_1, s_2, t), Y(s_1', s_2', t')) = \frac{1}{N} \sum_{i=1}^N [Y_i(s_1, s_2, t)-\bar{Y}(s_1, s_2, t)] [Y_i(s_1', s_2', t')-\bar{Y}(s_1', s_2', t')]$$

This matrix is apparently very large and would be difficult to either visualize or interpret. It doesn't make sense to flatten eventhing out and force space and time index onto the same grid. Therefore, I tried to simplify it to a marginal/conditional matrix. A few ways I can think of as follows: 


### Marginal correlation

First, I'll start with covariate between time points marginal over location: 

$$Cov(Y(t), Y(t')) = \frac{1}{N}\frac{1}{S_n} \sum_{i=1}^N \sum_{\mathbf{s}=1}^S [Y_i(\mathbf{s}, t)-\bar{Y}(., t)] [Y_i(\mathbf{s}, t')-\bar{Y}(., t')]$$

```{r}
marg_t_cov <- expand.grid(t1 = t, t2 = t)
marg_t_cov$cor <- NA

for(t1 in 1:nT){
  y1 <- df$Y[df$t==t[t1]]
  
  for(t2 in 1:nT){
    y2 <- df$Y[df$t==t[t2]]
    marg_t_cov[marg_t_cov$t1==t[t1]&marg_t_cov$t2==t[t2], "cor"] <- cor(y1, y2, method = "pearson")
    
  }
}

marg_t_cov %>%
  ggplot()+
  geom_tile(aes(x=t1, y=t2, fill=cor))+
  labs(title="Temporal correlation matrix marginal over location")

```


Use the same logic, we can calculate marginal correlation matrix over space. But I am not exactly sure who to visualize this matrix, since it is over four dimensions.  


```{r, eval=FALSE}
marg_s1_cov <- expand.grid(grid1=s1, grid2=s1)
marg_s1_cov$cor <- NA

for(sid1 in seq_along(s1)){
  y1 <- df$Y[df$s1==s1[sid1]]
  
  for(sid2 in seq_along(s1)){
    y2 <- df$Y[df$s1==s1[sid2]]
    marg_s1_cov[marg_s1_cov$grid1==s1[sid1]&marg_s1_cov$grid2==s1[sid2], "cor"] <- cor(y1, y2, method = "pearson")
    
  }
}

marg_s1_cov %>%
  ggplot()+
  geom_tile(aes(x=grid1, y=grid2, fill=cor))+
  labs(title = "Marginal correlation matrix along s1")
```

```{r, eval=FALSE}
marg_s2_cov <- expand.grid(grid1=s2, grid2=s2)
marg_s2_cov$cor <- NA

for(sid1 in seq_along(s2)){
  y1 <- df$Y[df$s2==s2[sid1]]
  
  for(sid2 in seq_along(s2)){
    y2 <- df$Y[df$s2==s2[sid2]]
    marg_s2_cov[marg_s2_cov$grid1==s2[sid1]&marg_s2_cov$grid2==s2[sid2], "cor"] <- cor(y1, y2, method = "pearson")
    
  }
}

marg_s2_cov %>%
  ggplot()+
  geom_tile(aes(x=grid1, y=grid2, fill=cor))+
  labs(title = "Marginal correlation matrix along s2")
```


### Multilevel model

Assume spatial and temporal effects can be factorized: 

$$Y_i(\mathbf{s}, t) = f_0(\mathbf{s}, t)+b_i(\mathbf{s}, t)=f_0(\mathbf{s}, t)+\sum_{k=1}^K\xi_{ik}\phi_{k}(\mathbf{s})+\sum_{l=1}^L\zeta_{il}\psi_{k}(t)$$

We can fit this model and visualize individual random effect over space or time. 

```{r}
# basis function for spatial random effect
phi_mat <- smoothCon(ti(s1, s2, k=3, bs = c("cr", "cr")), data= df %>% select(s1, s2) %>% distinct(.))
# dim(phi_mat[[1]]$X)
phi_mat <- data.frame(df %>% select(s1, s2) %>% distinct(.), phi_mat[[1]]$X)
colnames(phi_mat) <- c("s1", "s2", paste0("phi", 1:4))
```


```{r}
# basis function for temporal random effect
psi_mat <- smoothCon(s(t, k=4, bs = "cr"), data= data.frame(t))
# dim(psi_mat[[1]]$X)
psi_mat <- data.frame(t=t, psi_mat[[1]]$X)
colnames(psi_mat) <- c("t", paste0("psi", 1:4))
```


```{r}
df_fit <- df %>%
  select(-phi1, -phi2) %>%
  left_join(phi_mat, by = c("s1", "s2")) %>%
  left_join(psi_mat, by = "t")

fit_glmm <- bam(Y ~ ti(s1, s2, t, k=3, bs = c("cr", "cr", "cr"))+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re")+
                     s(id, by=psi1, bs="re")+
                     s(id, by=psi2, bs="re")+
                     s(id, by=psi3, bs="re")+
                     s(id, by=psi4, bs="re"), 
                data = df_fit,
                method = "fREML",
                discrete = TRUE)

# covariance
# length(fit_glmm$coefficients)
# dim(fit_glmm$Ve)
```


```{r}
# See the estimated functional effect for one subject
glmm_pred <- predict(fit_glmm, type = "terms")
# dim(glmm_pred)

pred_df <- df_fit %>% select(id, s1, s2, t) %>%
  mutate(mu_est = glmm_pred[, 1], 
         sp_est = apply(glmm_pred[, 2:5], 1, sum), 
         t_est = apply(glmm_pred[, 6:9], 1, sum))

rand_id <- sample(1:N, size = 4)

```


- Visualize random effect


```{r} 
# Spatial
pred_df %>%
  filter(id %in% rand_id) %>%
  select(id, s1, s2, sp_est) %>%
    ggplot()+
    geom_tile(aes(x=s1, y=s2, fill = sp_est))+
    facet_wrap(~id)+
  labs(title = "Individual spatial random effect")
  
```


```{r}
pred_df %>%
  filter(id %in% rand_id) %>%
  mutate(id = as.factor(id)) %>%
  select(id, t, t_est) %>%
    ggplot()+
    geom_line(aes(x=t, y=t_est, col = id, group = id))
  
```

- First of all, potentially, there can be a lot of basis functions for random effects
- This is not a "correlation matrix" but a function. It is also on the individual level, not population level. 
- Some potential ideas: semivariogram? 



# Some questions

1. Is it necessary to make the basis functions orthogonal/orthonormal? It doesn't seem to make sense considering time and space are such different measure. What does "time" and "space" being orthogonal mean? 