---
title: "Simulating spatial-temporal data"
author: "`r Sys.Date()`"
output: 
  html_document:
    self_contained: yes
    number_sections: true
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    font: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(1026)

library(mgcv)
library(nlme)
library(tidyverse)
library(knitr)
library(kableExtra)
library(mvtnorm)
library(ggpubr)
library(here)

source(here("Code/RandomField.R"))
```


# Generate from model 

The simulation, technically, would be established on a continuous space-time domain, even though the final "observations" would be a discrete realization.

Take a Gaussian process for example:

1. The observation is composed of a true latent process and an error process.

$$Y_i(\mathbf{s}, t) = \eta_i(\mathbf{s}, t) +\epsilon_i(\mathbf{s}, t)$$
2. The true latent process is composed of a fixed process and a random (subject-specific) process. 

$$\eta_i(\mathbf{s}, t) = \mu(\mathbf{s}, t)+b_i(\mathbf{s}, t)$$

- $\mu(\mathbf{s}, t)$ is the population mean function, shared across subjects
- $b_i(\mathbf{s}, t)$ is the individual-level random effect

3. The error process is spatially-correlated. Correlation is introduced through a moving average random field: 

$$\epsilon_i(\mathbf{s}, t) =  \frac{1}{N_r}\sum_{\mathbf{s'} \in S_r}Z(\mathbf{s'}, t)$$


where:

- $S_r$ is a neighborhood around $\mathbf{s}$ where the radius is r
- $N_r$ is the number of spacial points within neighborhood $S_r$
- $Z(\mathbf{s'}, t)$ is a white noise process


4. The second outcome is generated from the first outcome, and the correlation is time-varying

$$Y'_i(\mathbf{s},t) = \beta_i(t)Y_i(\mathbf{s},t) + b'_i(\mathbf{s},t) + \epsilon_i(\mathbf{s}, t)$$

# Example

```{r grid}
# generate a 2D image of 256 by 256
s1 <- s2 <- 1:256
t <- seq(0.2, 1, by = 0.2)
nT <- length(t)

```


## Moving average error 

```{r ma_error}
# generate white noise
Zmat <- matrix(rnorm(length(s1)*length(s2), 0, 1), length(s1), length(s2))

ma_k5 <- MA_rand_field(5, Zmat)
ma_k15 <- MA_rand_field(15, Zmat)
ma_k25 <- MA_rand_field(25, Zmat)

df_ma <- expand.grid(s1=s1, s2=s2) %>%
  mutate(original = as.vector(Zmat),
         k5=as.vector(ma_k5),
         k15=as.vector(ma_k15),
         k25=as.vector(ma_k25))

```


```{r, fig.height=6, fig.width=6}
# visualization
ggarrange(
  df_ma %>%
    ggplot()+
    geom_tile(aes(x=s1, y=s2, fill=original))+
    labs(title = "Original white noise"),
   
  df_ma %>%
    ggplot()+
    geom_tile(aes(x=s1, y=s2, fill=k5))+
    labs(title = "Moving average (k=5)"),

  
  df_ma %>%
    ggplot()+
    geom_tile(aes(x=s1, y=s2, fill=k15))+
    labs(title = "Moving average (k=15)"),

  
  df_ma %>%
    ggplot()+
    geom_tile(aes(x=s1, y=s2, fill=k25))+
    labs(title = "Moving average (k=25)"),
  
  nrow = 2, ncol = 2, common.legend = T

)
  
```

In this case the spatial correlation would have to do with the kernel size used to create the moving average field. It looks like the greater kernel size is, the stronger spatial correlation (with neighborhoods) would be.

For this report, I will set K = 15. 


## The mean process

A few things need clarification before a proper mean could be generated. For example:

- Is fixed effect time-varying? space-varying? What about random effects?
- Are coefficients time/space-varying?
- Does correlation exists in anywhere except for noise? 
- Do we need iid noise on top of the spatial correlated noise? 

For now, perhaps I should start with the simplest structure: 


Similar to the dynamic prediction project, I set the population mean as zero: $\mu(\mathbf{s}, t) = 0$, mostly because of **potential unidentifiability**. 

For random effect, I set up one random slope for (non-factorized) time, and one random slope for space. Specifically, the "spatial function" is the standardized distance away from a center $(c_1, c_2)$. 

PS Perhaps this center can also be subject-specific? Now I'm just gonna set $c_1=c_2=128.5$ for all subjects (midpoint of axis). 

\[\begin{aligned}
b_i(\mathbf{s}, t) &= \phi_1(\mathbf{s})\xi_{i1}+t\xi_{i2}, \ \xi_{ik} \sim N(0, \sigma^2_{k}) \\
\phi_1(\mathbf{s}) &= \sqrt{(\frac{s_1-c_1}{256})^2+(\frac{s_2-c_2}{256})^2}
\end{aligned}\]


## Correlation between two outcomes

\[\begin{aligned}
Y'_i(\mathbf{s},t) &= 2tY_i(\mathbf{s},t) + b'_i(\mathbf{s},t)+\sigma^2(\mathbf{s}, t)\\
b'_i(\mathbf{s}, t) &= \phi_1(\mathbf{s})\xi'_{i1}+t\xi'_{i2}, \ \xi'_{ik} \sim N(0, \sigma^2_{k})
\end{aligned}\]


## One-subject example

```{r gen_score}
K <- 2
# randoms core of Y1 and Y2
xi <- mvtnorm::rmvnorm(1, mean = rep(0, 2*K), sigma = diag(rep(1, 2*K)))
```


```{r}
df_ma$phi1 <- sqrt((df_ma$s1/256-mean(s1)/256)^2+(df_ma$s2/256-mean(s2)/256)^2)
df <- expand_grid(df_ma, t=t)

# the first outcome
df <- df %>%
  mutate(Y1 = phi1*xi[1]+t*xi[2]+k15)

# the second outcome
df <- df %>% mutate(Y2 = 2*t*Y1+xi[3]*phi1+xi[4]*t)
df$Y2 <- df$Y2+rnorm(nrow(df), 0, 0.2)

```

```{r, fig.height=7, fig.width=15}
ggarrange(
  
  df %>% ggplot()+geom_tile(aes(x=s1, y=s2, fill=Y1))+
    facet_wrap(~t, nrow = 1)+
    labs(y = "Y1", x = ""),
  
  df %>% ggplot()+geom_tile(aes(x=s1, y=s2, fill=Y2))+
    facet_wrap(~t, nrow = 1)+
    labs(y = "Y2", x = ""),
  
  ncol = 1, common.legend = T
)
```

```{r, fig.height=3, fig.width=15}
df %>% 
  ggplot(aes(x=Y1, y=Y2))+
  geom_point(size = 0.5)+
  geom_smooth(formula = 'y~x', method = "lm")+
  facet_wrap(~t, nrow = 1)+
  stat_cor(method = "pearson")
```

# Multi-subject example

Thus the full expression of data generation scheme:

\[\begin{aligned}

Y_i(\mathbf{s},t) &=\phi_1(\mathbf{s})\xi_{i1}+t\xi_{i2}+\epsilon_i(
\mathbf{s}, t)\\
Y'_i(\mathbf{s},t) &= 2tY_i(\mathbf{s},t)+\phi_1(\mathbf{s})\xi'_{i1}+t\xi'_{i2}+\sigma^2(\mathbf{s}, t)
\end{aligned}\]

- For simplicity, $\xi_{ik}, \xi'_{ik}$ are indenpendently generated from $N(0, 1)$
- $\sigma^2(\mathbf{s}, t)$ is a white noise generated from $N(0, 0.2^2)$
- $\epsilon_i$ is a moving average process of a 2D white noise field from $N(0, 1)$, derived by a 15 by 15 filter.

In the real dataset, the correlation would be different for each lesion. However, the current set up seems to imply constant correlation across subjects

```{r}
N <- 10 # sample size
# generate scores
xi_mat <- mvtnorm::rmvnorm(N, mean = rep(0, 2*K), sigma = diag(rep(1, 2*K)))

# container
df_list <- list()
```


```{r}
for(i in 1:N){
  # moving average noise
  Zmat <- matrix(rnorm(length(s1)*length(s2), 0, 1), length(s1), length(s2))
  MAmat <- MA_rand_field(15, Zmat)
  df_i <- expand.grid(s1=s1, s2=s2) %>% mutate(err =as.vector(MAmat))
  
  # first outcome
  df_i$phi1 <- sqrt((df_i$s1/256-mean(s1)/256)^2+(df_i$s2/256-mean(s2)/256)^2)
  df_i <- expand_grid(df_i, t=t)
  df_i <- df_i %>%
    mutate(Y1 = phi1*xi_mat[i,1]+t*xi_mat[i,2]+err,
           Y2 = 2*t*Y1+xi_mat[i,3]*phi1+xi_mat[i,4]*t)
  df_i$Y2 <- df_i$Y2+rnorm(nrow(df_i), 0, 0.2)
  
  df_list[[i]] <- df_i
  
}
```


```{r cor_table}
lapply(df_list, function(x){x %>% group_by(t) %>%
    summarize(cor = cor(Y1, Y2))}) %>%
  bind_rows(.,.id = "subject") %>% 
  pivot_wider(., names_from = t, values_from = cor) %>% 
  kable(caption = "Correlation", digits = 3) %>%
  kable_styling(full_width = F)
```

- Below is an example of subject 9, whose correlation increased cross zero

```{r, fig.height=10, fig.width=15}
ggarrange(
  df_list[[9]] %>% ggplot()+
    geom_tile(aes(x=s1, y=s2, fill=Y1))+
    facet_wrap(~t, nrow = 1)+
    labs(y = "Y1", x = ""),
  
  df_list[[9]] %>% ggplot()+
    geom_tile(aes(x=s1, y=s2, fill=Y2))+
    facet_wrap(~t, nrow = 1)+
    labs(y = "Y2", x = ""),
  
  df_list[[9]] %>% 
    ggplot(aes(x=Y1, y=Y2))+
    geom_point(size = 0.5)+
    geom_smooth(formula = 'y~x', method = "lm")+
    facet_wrap(~t, nrow = 1)+
    stat_cor(method = "pearson"),
  
  ncol = 1, common.legend = T
)
```


- And below is an example of subject one, whose correlation stayed positive


```{r, fig.height=10, fig.width=15}
ggarrange(
  df_list[[1]] %>% ggplot()+
    geom_tile(aes(x=s1, y=s2, fill=Y1))+
    facet_wrap(~t, nrow = 1)+
    labs(y = "Y1", x = ""),
  
  df_list[[1]] %>% ggplot()+
    geom_tile(aes(x=s1, y=s2, fill=Y2))+
    facet_wrap(~t, nrow = 1)+
    labs(y = "Y2", x = ""),
  
  df_list[[1]] %>% 
    ggplot(aes(x=Y1, y=Y2))+
    geom_point(size = 0.5)+
    geom_smooth(formula = 'y~x', method = "lm")+
    facet_wrap(~t, nrow = 1)+
    stat_cor(method = "pearson"),
  
  ncol = 1, common.legend = T
)
```